Extended Data

### 2022-03-29 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd
from google doc
minor proof-reading edits

Sean McMahon^9,11^ (Orcid ID : 0000-0001-8302-6908)

@@

###~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
## some stats/ 
species_table <- read.csv(file = "tables_figures/species_table.csv")
n_dendro_trees <- sum(na.omit(species_table[, "n.trees"]))
n_dendro_trees_SCBI <- 119 #doing this the lazy way. summed from species_table rather than coded. 
n_dendro_trees_HF <- 321 #doing this the lazy way. summed from species_table rather than coded. 
n_tree_years <- sum(na.omit(species_table[, "n.tree.years"]))
n_cores_focal_sites <- sum(na.omit(species_table[, "n.cores"]))
chronology_table <- read.csv(file = "tables_figures/chronology_table.csv")
n_chronologies <- nrow(chronology_table)  
n_cores_sites <- length(unique(chronology_table[, "Location"]))
n_cores_species <- length(unique(chronology_table[, "Species"]))
n_chronologies_SCBI <- 12
n_chronologies_HF <- 4
```

@@

![**Figure 1 | Summary of temperate deciduous tree growth responses to warmer spring temperatures.** (a) Schematic illustrating parameters of interest and summarizing predominant responses of each to warmer maximum temperatures during a 'critical temperature window', defined as the period with the strongest temperature control over *DOY~25~*; (b) Variable definitions and summary of responses to warmer spring temperatures at two temperate forests -- Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest -- and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate increases and decreases that are significant both statistically (p<0.05) and biologically (effect size >3%$^\circ$C^-1^), '-' indicates no significant correlation, and 'mixed' indicates a mix of significant and non-significant correlations, often in different directions.](tables_figures/schematic_summary_with_table.png){ width=70% }

@@

Both canopy foliage phenology and the timing of stem growth shifted earlier as spring temperatures increased (Fig. `r DOY_timing`, `r EDF_pheno_all`). 
We found a consistent effect of temperature (*T~max~* or *T~min~*) throughout the spring, but the strongest effects on stem growth timing were found using *T~max~* during a critical temperature window (CTW). 
CTW was identified by measuring the correlation between weekly *T~max~* and *DOY~25~* for all combinations of consecutive weeks from January 1 to mean *DOY~25~* for each xylem architecture-site combination (`r EDF_climwin`).
The CTW was defined as the weeks for which this correlation was strongest, and mean *T~max~* during this period (CTW *T~max~*) used as our independent variable.

@@

For ring- and diffuse- porous species at both sites, warmer CTW *T~max~* resulted in earlier achievement of seasonal milestones.
Consistent with findings from previous studies<!--[@parmesan_globally_2003;--><!--@ibanez_forecasting_2010;--><!--@friedl_tale_2014]-->, leaf phenological milestones advanced at both sites (Fig. `r DOY_timing`a-b, `r EDT_phenology_parameters`), with greenup (DOY when EVI2 first crossed 15% of the segment EVI2 amplitude) advancing 3.45 days/$^\circ$C at SCBI (p=0.016) and 2.4 days/ $^\circ$C at Harvard Forest (p=0.1).
Similarly, with the exception of ring porous *DOY~75~* at SCBI, the stem growth milestones *DOY~25~*, *DOY~50~*, and *DOY~75~* decreased with CTW *T~max~* (Figs. `r schematic_summary`, `r DOY_timing`c-d; `r EDF_pheno_all`).
Specifically, *DOY~25~*, *DOY~50~*, and *DOY~75~* advanced 0-1.7 days/ $^\circ$C for ring porous species and 2.8-2.9 days/ $^\circ$C for diffuse porous species at SCBI, and 10.3-12.3 days/ $^\circ$C for ring porous species and 0.9-4.2 days/ $^\circ$C for diffuse porous species at Harvard Forest (`r EDT_phenology_parameters`).

@@

In contrast to the pronounced effects of CTW *T~max~* on the timing of growth, its effects on *g~max~* and $\Delta DBH$ were usually weak and inconsistent (Fig. `r schematic_summary`, `r EDF_pheno_all`). 
Specifically, *g~max~*, which occurred on average within 5 days of *DOY~50~*, displayed extremely small negative changes (Harvard Forest) or changes in opposite directions (SCBI) in relationship to CTW *T~max~* for ring- and diffuse- porous species.
$\Delta DBH$ displayed a weak positive relationship (+0.0029-0.008 cm/ $^\circ$C) with CTW *T~max~* at SCBI and a weak negative relationship (-0.02-0.003 cm/ $^\circ$C) at Harvard Forest (`r EDF_pheno_Tmax_climwin`).

@@

![**Figure 3 | Sensitivity of annual growth, as derived from tree-rings, to monthly mean maximum temperatures (*T~max~*), for `r n_chronologies` chronologies from `r n_cores_sites` sites across eastern North America** (`r EDF_map `). Colors indicate the correlation between monthly *T~max~* and a dimensionless ring width index (RWI) derived from the multiple trees that form each chronology and emphasizing interannual variability associated with climate. Chronologies are grouped by xylem porosity and ordered by mean April *T~max~*. Plots are annotated to highlight records from our two focal sites, the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest (HF) (`r EDT_species`). Species analyzed and numbers of significant correlations to *T~max~* are summarized in `r EDT_tree_ring_summary`, and chronology details are given in `r chron_table`.](tables_figures/tree_rings_tmax.png){ width=100% }

@@

Cameron Dow and Kristina J. Anderson-Teixeira conceived the ideas and designed the study; 
Cameron Dow, Loïc D’Orangeville, Erika B. Gonzalez-Akre, Ryan Helcoski, Grant L. Harley, Justin T. Maxwell, Ian R. McGregor, William McShea, Neil Pederson, Alan J. Tepley, and Kristina J. Anderson-Teixeira collected or oversaw collection of data; 
Cameron Dow, Albert Y. Kim, Valentine Herrmann, Justin T. Maxwell, Ian R. McGregor, Sean McMahon analyzed the data or provided analytical tools; 
Cameron Dow and Kristina J. Anderson-Teixeira led the writing of the manuscript. 
All authors contributed critically to the drafts and gave final approval for publication.

>>>>

Sean M. McMahon^9,11^ (Orcid ID : 0000-0001-8302-6908)

@@

###~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
## some stats/ 
species_table <- read.csv(file = "tables_figures/species_table.csv")
n_dendro_trees <- sum(na.omit(species_table[, "n.trees"]))
n_dendro_trees_SCBI <- 119 #doing this the lazy way. summed from species_table rather than coded. 
n_dendro_trees_HF <- 321 #doing this the lazy way. summed from species_table rather than coded. 
n_tree_years <- sum(na.omit(species_table[, "n.tree.years"]))
n_cores_focal_sites <- sum(na.omit(species_table[, "n.cores"]))
chronology_table <- read.csv(file = "tables_figures/chronology_table.csv")
n_chronologies <- nrow(chronology_table)  
n_cores_sites <- length(unique(chronology_table[, "Location"]))
n_cores_species <- length(unique(chronology_table[, "Species"])) - 1 #the -1 accounts for the Quercus spp. record (combination of white oak group)
n_chronologies_SCBI <- 12
n_chronologies_HF <- 4
```

@@

![**Figure 1 | Summary of temperate deciduous tree growth responses to warmer spring temperatures.** (a) Schematic illustrating parameters of interest and summarizing predominant responses of each to warmer maximum temperatures during a 'critical temperature window', defined as the period with the strongest temperature control over *DOY~25~*; (b) Variable definitions and summary of responses to warmer spring temperatures at two temperate forests -- Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest -- and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate increases and decreases that are significant both statistically (p<0.05) and biologically (effect size >3%/ $^\circ$C), '-' indicates no significant correlation, and 'mixed' indicates a mix of significant and non-significant correlations, often in different directions.](tables_figures/schematic_summary_with_table.png){ width=70% }

@@

Both canopy foliage phenology and the timing of stem growth shifted earlier as spring temperatures increased (Fig. `r DOY_timing`, `r EDF_pheno_all`). 
We found a consistent effect of temperature (*T~max~* or *T~min~*) throughout the spring, but the strongest effects on stem growth timing were found using *T~max~* during a critical temperature window (CTW). 
CTW was identified by assessing the correlation between weekly *T~max~* and *DOY~25~* for all combinations of consecutive weeks from January 1 to mean *DOY~25~* for each xylem architecture-site combination (`r EDF_climwin`).
The CTW was defined as the weeks for which this correlation was strongest, and mean *T~max~* during this period (CTW *T~max~*) used as our independent variable.

@@

For ring- and diffuse- porous species at both sites, warmer CTW *T~max~* resulted in earlier achievement of seasonal milestones.
Consistent with findings from previous studies<!--[@parmesan_globally_2003;--><!--@ibanez_forecasting_2010;--><!--@friedl_tale_2014]-->, leaf phenological milestones advanced at both sites (Fig. `r DOY_timing`a-b, `r EDT_phenology_parameters`), with greenup (DOY when EVI2 first crossed 15% of the segment EVI2 amplitude) advancing 3.5 days/$^\circ$C at SCBI (p=0.016) and 2.4 days/ $^\circ$C at Harvard Forest (p=0.1).
Similarly, with the exception of ring porous *DOY~75~* at SCBI, the stem growth milestones *DOY~25~*, *DOY~50~*, and *DOY~75~* decreased with CTW *T~max~* (Figs. `r schematic_summary`, `r DOY_timing`c-d; `r EDF_pheno_all`).
Specifically, *DOY~25~*, *DOY~50~*, and *DOY~75~* advanced 0-1.7 days/ $^\circ$C for ring porous species and 2.8-2.9 days/ $^\circ$C for diffuse porous species at SCBI, and 10.3-12.3 days/ $^\circ$C for ring porous species and 0.9-4.2 days/ $^\circ$C for diffuse porous species at Harvard Forest (`r EDT_phenology_parameters`).

@@

In contrast to the pronounced effects of CTW *T~max~* on the timing of growth, its effects on *g~max~* and $\Delta DBH$ were usually weak and inconsistent (Fig. `r schematic_summary`, `r EDF_pheno_all`). 
Specifically, *g~max~*, which occurred on average within 5 days of *DOY~50~*, displayed extremely small negative changes (Harvard Forest) or changes in opposite directions (SCBI) in relationship to CTW *T~max~* for ring- and diffuse- porous species.
$\Delta DBH$ displayed a weak positive relationship (+0.003-0.008 cm/ $^\circ$C) with CTW *T~max~* at SCBI and a weak negative relationship (-0.02-0.003 cm/ $^\circ$C) at Harvard Forest (`r EDF_pheno_Tmax_climwin`).

@@

![**Figure 3 | Sensitivity of annual growth, as derived from tree-rings, to monthly mean maximum temperatures (*T~max~*), for `r n_chronologies` chronologies from `r n_cores_sites` sites across eastern North America**. Colors indicate the correlation between monthly *T~max~* and a dimensionless ring width index (RWI) derived from the multiple trees that form each chronology and emphasizing interannual variability associated with climate. Chronologies are grouped by xylem porosity and ordered by mean April *T~max~*. Plots are annotated to highlight records from our two focal sites, the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest (HF; `r EDT_species`). Sites included are mapped in `r EDF_map `, species analyzed and numbers of significant correlations to *T~max~* are summarized in `r EDT_tree_ring_summary`, and chronology details are given in `r chron_table`.](tables_figures/tree_rings_tmax.png){ width=100% }

@@

Cameron Dow and Kristina J. Anderson-Teixeira conceived the ideas and designed the study; 
Cameron Dow, Loïc D’Orangeville, Erika B. Gonzalez-Akre, Ryan Helcoski, Grant L. Harley, Justin T. Maxwell, Ian R. McGregor, William McShea, Neil Pederson, Alan J. Tepley, and Kristina J. Anderson-Teixeira collected or oversaw collection of data; 
Cameron Dow, Albert Y. Kim, Valentine Herrmann, Justin T. Maxwell, Ian R. McGregor, Sean M. McMahon analyzed the data or provided analytical tools; 
Cameron Dow and Kristina J. Anderson-Teixeira led the writing of the manuscript. 
All authors contributed critically to the drafts and gave final approval for publication.
